\name{}: machine learning over interaction networks and ontology-based background knowledge predicts drug--target interactions

### 2021-04-22 paper/main.tex
.

Potential biases resulting from
the underlying datasets \citep{Pahikkala2014} which may affect model
evaluation and comparison pose a challenge for DTI
prediction. Firstly, novel drugs are often developed by
altering non-functional components of a drug, leading to two and more
very similar drugs designed to target the same proteins
\cite{Overington2006}. This can result in a bias when it leads to
hidden duplicates or highly similar compounds that are distributed
among training and evaluation dataset, resulting in a better
(measured) predictive performance than it would be expected when the
model is applied to identify drugs that target a protein for which no
drugs yet exist. Secondly, some proteins (which we call \textit{hub
  proteins}) have significantly more known interactions with drugs
than others. In the STITCH database, $5\%$ of the proteins have $40\%$
of the interactions, and similar distributions are present in other
datasets \cite{Drugbank2007, Drugbank2017}; preferentially predicting
these proteins may increase predictive performance while not
reflecting the actual performance when applied to a new protein (i.e.,
a protein for which no interactions are known). These differences in
the number of drugs targeting certain proteins may be the result of
study bias where more ``valuable'' proteins have more drugs designed
to target them due to their involvement in more common diseases (or
diseases for which drugs can be more profitably marketed). This might
affect common evaluation schemes \cite{Survey2018} where it is possible
to exploit these biases within DTI prediction \cite{}\todo[inline]{cite or delete}.
\citet{vanLaarhoven2014} showed that several bias can be exploited on
the dataset of \citet{Yamanishi2008}.

@@

As the difference in performance with different splitting schemes is
quite large, we further evaluated additional drug--target interaction
and drug--target affinity prediction methods that were trained and
evaluation on other datasets.  Following the results of MolTrans
\cite{MolTrans2020}, we reevaluated DeepDTI \cite{DeepDTI2017},
DeepDTA \cite{DeepDTA2018}, DeepConv-DTI \cite{DeepConvDTI2019}, and
MolTrans itself on the BioSnap dataset \cite{BioSnap2018} and compared it to our
``na\"ive'' predictor as well as \name{} (see Table
\ref{tab:comparison2}).  MolTrans was evaluated over the drug--target
pair and the protein split; we were able to reproduce the MolTrans
results (Table \ref{tab:comparison2}), showing a substantial
difference based on the splitting scheme. We additionally computed the
$MicroAUC_p$ score for all considered methods, leading to similar 
results as observed on the Yaminishi dataset. As \name{} considers less drugs 
than alternative approaches, we perform a two-sided T-test yielding $p<0.0001$ for 
$\alpha<0.01$ showing the significance of \name's performance gain over other state-of-the-art methods.\todo[inline]{Formulation correct? Confidence interval?}

>>>>

Potential biases resulting from the underlying datasets
\citep{Pahikkala2014} which may affect model evaluation and comparison
pose a challenge for DTI prediction. Firstly, novel drugs are often
developed by altering non-functional components of a drug, leading to
two and more very similar drugs designed to target the same proteins
\cite{Overington2006}. This can result in a bias when it leads to
hidden duplicates or highly similar compounds that are distributed
among training and evaluation dataset, resulting in a better
(measured) predictive performance than it would be expected when the
model is applied to identify drugs that target a protein for which no
drugs yet exist. Secondly, some proteins (which we call \textit{hub
  proteins}) have significantly more known interactions with drugs
than others. In the STITCH database, $5\%$ of the proteins have $40\%$
of the interactions, and similar distributions are present in other
datasets \cite{Drugbank2007, Drugbank2017}; preferentially predicting
these proteins may increase predictive performance while not
reflecting the actual performance when applied to a new protein (i.e.,
a protein for which no interactions are known). These differences in
the number of drugs targeting certain proteins may be the result of
study bias where more ``valuable'' proteins have more drugs designed
to target them due to their involvement in more common diseases (or
diseases for which drugs can be more profitably marketed). This might
affect common evaluation schemes where it is possible to exploit these
biases within DTI prediction \cite{Survey2018}.
\citet{vanLaarhoven2014} showed that several bias can be exploited on
the dataset of \citet{Yamanishi2008}.

@@

As the difference in performance with different splitting schemes is
quite large, we further evaluated additional drug--target interaction
and drug--target affinity prediction methods that were trained and
evaluation on other datasets.  Following the results of MolTrans
\cite{MolTrans2020}, we reevaluated DeepDTI \cite{DeepDTI2017},
DeepDTA \cite{DeepDTA2018}, DeepConv-DTI \cite{DeepConvDTI2019}, and
MolTrans itself on the BioSnap dataset \cite{BioSnap2018} and compared it to our
``na\"ive'' predictor as well as \name{} (see Table
\ref{tab:comparison2}).  MolTrans was evaluated over the drug--target
pair and the protein split; we were able to reproduce the MolTrans
results (Table \ref{tab:comparison2}), showing a substantial
difference based on the splitting scheme. We additionally computed the
$MicroAUC_p$ score for all considered methods, leading to similar 
results as observed on the Yaminishi dataset. % As \name{} considers less drugs 
% than alternative approaches, we perform a two-sided T-test yielding $p<0.0001$ for 
% $\alpha<0.01$ showing the significance of \name's performance gain over other state-of-the-art methods.\todo[inline]{Formulation correct? Confidence interval?}
